Elephant Seal Age JAE Intraspecific Variation

### 2024-05-23 paper/paperv3.qmd
raw data and surv/recr figs

---
title: "Elephant Seal Age JAE Intraspecific Variation"
format: 
  html: 
    echo: false
    fig-width: 4.5
    fig-height: 4.5
    code-fold: true
    code-tools: true
    warning: false
editor: visual
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(ggridges)
library(patchwork)
source(here::here("R/format.R"))

@@

age_dist <- sealdat %>% 
  filter(age < 22) %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "grey50") + 
  labs(x = "Age (Years)", y = "Number of individuals") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

age_dist

@@

```{r}
#| label: fig-breeding-mod 
#| fig-width: 9
#| fig-height: 6

@@

# Figure with raw data and model predictions
breed_fig <- ggplot(breed_pred, aes(x = age)) +
  # Prediction across years (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Predictions for specific years (random effect)
  geom_line(aes(y = predicted, group = interaction(year_fct, age_class)),
            breed_pred_year,
            linewidth = 0.75, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) + 
  # Year labels
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 20),
            x = 20.2,
            hjust = 0,
            size = 3) +
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 4),
            x = 3.8,
            hjust = 1,
            size = 3) +
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_x_continuous("Age", limits = c(3.5, 20.5)) +
  scale_y_continuous("Breeding %", 
                     limits = c(NA, 1), 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))

# Coefficient panel
breed_coef_fig <- breed_summ %>% 
  mutate(model = fct_rev(model)) %>% 
  ggplot(aes(color = model)) + 
  geom_pointrange(aes(ymin = age_coef - age_coef_se,
                      y = age_coef,
                      ymax = age_coef + age_coef_se,
                      x = model)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  labs(y = "Age coefficient") +
  theme_few() +
  theme(legend.position = "none",
        axis.title.x = element_blank())

(breed_fig | breed_coef_fig) + plot_layout(widths = c(3, 1))

@@

# Fit pup *survival* models pre- and post-senescence
pup_surv_model1 <- glmer(
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age < age_senesce),
  family = "binomial"
)
pup_surv_model2 <- glmer(
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age >= age_senesce),
  family = "binomial"
)

@@

# Summary table
tibble(
  response = rep(c("Survival", "Recruitment"), each = 2),
  n = map_dbl(
    list(pup_surv_model1@frame, 
         pup_surv_model2@frame, 
         pup_recr_model1@frame, 
         pup_recr_model2@frame),
    nrow
  ),
  model = rep(c("Pre-senescence", "Post-senescence"), 2),
  age_coef = c(fixef(pup_surv_model1)["age10"],
               fixef(pup_surv_model2)["age10"],
               fixef(pup_recr_model1)["age10"],
               fixef(pup_recr_model2)["age10"]),
  age_coef_se = c(summary(pup_surv_model1)$coef["age10", "Std. Error"],
                  summary(pup_surv_model2)$coef["age10", "Std. Error"],
                  summary(pup_recr_model1)$coef["age10", "Std. Error"],
                  summary(pup_recr_model2)$coef["age10", "Std. Error"]),
  `p(coef<0)` = c(pnorm(summary(pup_surv_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_surv_model2)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model2)$coef["age10", "z value"]))
) %>% 
  mutate(across(where(is.numeric), \(x) signif(x, 3)))

@@

```{r}
#| label: fig-repro-age
#| fig-cap: "Pup survival and recruitment are not significantly affected by maternal age. Points are mean values for maternal age. Error bars show SE."
#| fig-width: 5.5

@@

repro_raw <- surv_recruit_data %>% 
  mutate(pupsex = factor(pupsex, labels = c("Female", "Male"))) %>% 
  group_by(age, pupsex) %>% 
  summarize(n_born = n(), 
            n_surv = sum(pup_survived_int),
            perc_surv = n_surv / n_born, 
            se_surv = (perc_surv * (1 - perc_surv) / n_born)^0.5,
            n_recr = sum(puprecruited),
            perc_recr = n_recr / n_born,
            se_recr = (perc_recr * (1 - perc_recr) / n_born)^0.5,
            .groups = "drop")

ggplot(surv_pred, aes(x = age)) +
  # Survival by age (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Raw survival and recruitment data
  geom_pointrange(aes(y = perc_surv,
                      ymin = perc_surv - se_surv,
                      ymax = perc_surv + se_surv),
                  repro_raw) +
  geom_pointrange(aes(y = perc_recr,
                      ymin = perc_recr - se_recr,
                      ymax = perc_recr + se_recr),
                  shape = 21,
                  filter(repro_raw, pupsex == "Female")) +
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_x_continuous("Age", limits = c(3.5, 18.5)) +
  scale_y_continuous("Pup survival/recruitment", 
                     limits = c(0, 0.65), 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  facet_wrap(~pupsex) +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))

>>>>

---
title: "Elephant Seal Age JAE Intraspecific Variation"
format: 
  html: 
    echo: false
    fig-width: 5
    fig-height: 5
    code-fold: true
    code-tools: true
    warning: false
editor: visual
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(ggridges)
library(patchwork)
source(here::here("R/format.R"))
theme_set(theme_few(base_size = 13))

@@

sighting_img <- sealdat %>% 
  group_by(animalID) %>% 
  summarize(yearborn = yearborn[1],
            first_obs = min(year),
            last_obs = max(year),
            .groups = "keep") %>% 
  reframe(yearborn = yearborn,
          year = yearborn:last_obs,
          age = year - yearborn,
          prerecruit = year < first_obs) %>% 
  filter(year > 1987) %>% 
  left_join(sealdat %>% 
              group_by(animalID, year) %>% 
              summarize(observed_int = max(observed_int),
                        .groups = "drop"),
            by = c("animalID", "year")) %>% 
  mutate(
    observed = case_when(
      age == 0 ~ "Born",
      prerecruit ~ "Pre-recruit",
      observed_int == 0 ~ "Non-breeder",
      observed_int == 1 ~ "Breeder",
      is.na(observed_int) ~ "Unobserved"
    ),
    observed = factor(observed, levels = c("Born",
                                           "Pre-recruit",
                                           "Non-breeder",
                                           "Breeder",
                                           "Unobserved"))
  ) %>% 
  group_by(animalID) %>% 
  mutate(longevity = max(age),
         yearborn = na.omit(yearborn)[1]) %>% 
  ungroup() %>% 
  arrange(yearborn, longevity, animalID) %>% 
  mutate(animalID = factor(animalID, levels = unique(animalID))) %>% 
  ggplot(aes(year, animalID)) +
  geom_raster(aes(fill = observed)) +
  scale_fill_viridis_d(direction = -1) +
  guides(fill = guide_legend(position = "inside", 
                             direction = "horizontal",
                             ncol = 1,
                             byrow = TRUE)) +
  labs(y ="Individual") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        legend.position.inside = c(0.01, 0.99),
        legend.justification = c(0, 1),
        legend.background = element_blank(),
        legend.key.size = unit(4, "mm"),
        legend.text = element_text(size = 8))

age_dist <- sealdat %>% 
  filter(age < 22) %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "grey50") + 
  labs(x = "Age (Years)", y = "Number of individuals") + 
  theme()

(sighting_img | age_dist) + plot_layout(widths = c(2, 1))

@@

```{r}
#| label: fig-breeding-mod 
#| fig-width: 7.5

@@

# Figure with raw data and model predictions
breed_fig <- ggplot(breed_pred, aes(x = age)) +
  # Prediction across years (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Predictions for specific years (random effect)
  geom_line(aes(y = predicted, group = interaction(year_fct, age_class)),
            breed_pred_year,
            linewidth = 0.75, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) + 
  # Year labels
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 20),
            x = 20.2,
            hjust = 0,
            size = 2.5) +
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 4),
            x = 3.8,
            hjust = 1,
            size = 2.5) +
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_x_continuous("Age", limits = c(3.5, 20.5)) +
  scale_y_continuous("Adult female breeding", 
                     limits = c(NA, 1), 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))

# Coefficient panel
breed_coef_fig <- breed_summ %>% 
  mutate(model = str_extract(model, ".*-"),
         model = fct_rev(model)) %>% 
  ggplot(aes(color = model)) + 
  geom_pointrange(aes(ymin = age_coef - age_coef_se,
                      y = age_coef,
                      ymax = age_coef + age_coef_se,
                      x = model)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  labs(x = "Senescence",
       y = "Age coefficient") +
  theme_few() +
  theme(legend.position = "none")

(breed_fig | breed_coef_fig) + 
  plot_layout(widths = c(3, 1)) +
  plot_annotation(tag_levels = "A")

@@

# Fit pup *survival* models pre- and post-senescence
# Commented-out formulae vary age10 coef by pupsex, end up with similar
# results. Can add to supplemental?
pup_surv_model1 <- glmer(
  # pup_survived_int ~ age10 : pupsex + (1 | animalID) + (1 | year),
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age < age_senesce),
  family = "binomial"
)
pup_surv_model2 <- glmer(
  # pup_survived_int ~ age10 : pupsex + (1 | animalID) + (1 | year),
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age >= age_senesce),
  family = "binomial"
)

@@

# Summary table
surv_recr_summ <- tibble(
  response = rep(c("Survival", "Recruitment"), each = 2),
  n = map_dbl(
    list(pup_surv_model1@frame, 
         pup_surv_model2@frame, 
         pup_recr_model1@frame, 
         pup_recr_model2@frame),
    nrow
  ),
  model = rep(c("Pre-senescence", "Post-senescence"), 2),
  age_coef = c(fixef(pup_surv_model1)["age10"],
               fixef(pup_surv_model2)["age10"],
               fixef(pup_recr_model1)["age10"],
               fixef(pup_recr_model2)["age10"]),
  age_coef_se = c(summary(pup_surv_model1)$coef["age10", "Std. Error"],
                  summary(pup_surv_model2)$coef["age10", "Std. Error"],
                  summary(pup_recr_model1)$coef["age10", "Std. Error"],
                  summary(pup_recr_model2)$coef["age10", "Std. Error"]),
  `p(coef<0)` = c(pnorm(summary(pup_surv_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_surv_model2)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model2)$coef["age10", "z value"]))
)

surv_recr_summ %>% 
  mutate(across(where(is.numeric), \(x) signif(x, 3))) %>% 
  knitr::kable()

@@

```{r}
#| label: fig-repro-age
#| fig-cap: "Offspring survival decreases as mothers senesce. **A** Survival of female and male offspring (filled points) is not negatively impacted by maternal age (blue line) prior to age 12 (vertical dashed line). After age 12, offspring survival declines with maternal age (red line), i.e., maternal effect senescence. Hollow points represent female offspring recruitment, which did not significantly decline with maternal age after age 12. **B** The coefficient of age w.r.t. offspring survival (solid points) was not significantly negative before age 12 (blue), but was significantly negative after (red). The coefficient of age w.r.t. female offspring recruitment (hollow points) was not significant in either maternal age class. Lines represent standard error of the coefficient, which is ommitted for the post-age 12 offspring recruitment coefficent (hollow red point) because the standard error was too large (14.8) to visually represent without hiding other patterns. Numbers at top indicate sample size in each model."
#| fig-width: 5.5

@@

repro_raw <- surv_recruit_data %>% 
  # Uncomment to group by female and male offspring
  # mutate(
  #   pupsex = factor(pupsex, 
  #                   labels = c("Female offspring", "Male offspring"))
  # ) %>% 
  # group_by(age, pupsex) %>% 
  group_by(age) %>% 
  summarize(n_born = n(), 
            n_surv = sum(pup_survived_int),
            perc_surv = n_surv / n_born, 
            se_surv = (perc_surv * (1 - perc_surv) / n_born)^0.5,
            f_born = sum(pupsex == "F"),
            f_recr = sum(puprecruited & pupsex == "F"),
            perc_recr = f_recr / f_born,
            se_recr = (perc_recr * (1 - perc_recr) / f_born)^0.5,
            .groups = "drop")

surv_recr_fig <- ggplot(surv_pred, aes(x = age)) +
  # Survival by age (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Raw survival and recruitment data
  geom_pointrange(aes(y = perc_surv,
                      ymin = perc_surv - se_surv,
                      ymax = perc_surv + se_surv),
                  repro_raw) +
  geom_pointrange(aes(y = perc_recr,
                      ymin = perc_recr - se_recr,
                      ymax = perc_recr + se_recr),
                  repro_raw,
                  shape = 21) +
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_x_continuous("Age", limits = c(3.5, 18.5)) +
  scale_y_continuous("Offspring survival/recruitment", 
                     limits = c(0, 0.65), 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  # Uncomment to facet by offspring sex
  # facet_wrap(~pupsex) +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))

# Survival/recruitment figure. In the caption, mention the post-senescence
# recruitment std err was omitted because it was 14.8, so it swamps all the
# other std errs
surv_recr_coef_fig <- surv_recr_summ %>% 
  mutate(response = fct_rev(str_sub(response, 1, 4)),
         model = fct_rev(str_extract(model, ".*-")),
         # Remove post-senescence recruitment std err
         age_coef_se = c(age_coef_se[1:3], 0)) %>% 
  ggplot(aes(x = response, ymax = 1.1)) +
  geom_pointrange(aes(ymin = age_coef - age_coef_se, 
                      y = age_coef,
                      ymax = age_coef + age_coef_se,
                      color = model,
                      shape = response),
                  position = position_dodge(width = 0.9)) +
  geom_text(aes(label = n, group = model), 
            y = 0.9, 
            size = 3,
            angle = 30,
            hjust = 0,
            position = position_dodge(width = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_shape_manual(values = c(16, 1)) +
  labs(x = "Response", y = "Age coefficient") +
  expand_limits(y = 1.1) +
  theme(legend.position = "none")

(surv_recr_fig | surv_recr_coef_fig) + 
  plot_layout(widths = c(3, 1)) +
  plot_annotation(tag_levels = "A")
